Review: Redefining Domestication: Adaptive Introgression during Crop Expansion

### 2017-06-13 paper.tex
edits to intro.

The process of domestication was once thought to be rapid and geographically constrained, with crops originating from a wild progenitor within one or more geographically defined centers followed by expansion to the modern-day extent of cultivation. \mbh{will need citation for this}
However, archaeological and genetic evidence are beginning to reveal that, in many cases, domestication has been temporally protracted and more diffuse \cite{brown2009complex, Meyer2016, wang2017, zhou2017}. \mbh{citations here to Li's paper, Purugganan rice paper, Brandon Gaut's recent grape paper} \lwang{I have added these three papers.}
This new conception of domestication emphasizes the role of beneficial gene flow (\emph{i.e.}, adaptive introgression) from locally adapted wild relatives during crop expansion after initial domestication.

Adaptive introgression has three components: hybridization between two genomes, backcrossing to one of the parents, and selection on different recombinant genotypes with progressively diminished linkage drag \cite{barton2001role, Feuillet200824}.
In domesticated species, adaptive introgression would consist of crop/wild hybrids backcrossing to a crop, retention and increase in frequency of adaptive wild haplotypes in the crop, and removal of undesirable wild background.
To date, literature on crop-wild gene flow has focused on the risk of introgression of transgenes from domesticated crops into wild relatives (for a review, \cite{stewart2003transgene}) and on modern plant breeding efforts to introgress desired traits from wild relatives (for a review, \cite{zamir2001improving, tanksley1997seed, hajjar2007use}). \mbh{cite one paper here...the best, most recent and comprehensive review}
The history of natural introgression of wild alleles into domesticated crops over evolutionary timescales has received considerably less attention.
However, recent tools and methods have been employed to detect genome-wide patterns of introgression, granting new insights into the prevalence of adaptive introgression in crop histories.
Emerging results suggest a need for reevaluation of the existing domestication paradigm.

@@

The decreasing cost of genome-wide resequencing and availability of reduced-representation genotyping (\emph{e.g.}, GBS and RAD-Seq), combined with new analytical methods, has facilitated comprehensive study of introgression across a number of species (\textbf{Table 1}).
High-density marker data can be used with haplotype-based and other methods to assign specific genomic regions to a taxon of origin and identify introgression across taxa \cite{Martin2015,Price2009,Lawson2012,pease2015,rosenzweig2016,geneva2015}.
The methods reviewed here do not include those marginally estimating introgression\slash migration rate as a component of demographic history (\emph{e.g.}, Approximate Bayesian Computation (ABC) \cite{beaumont2002}, diffusion approximations for demographic inference ($\delta a\delta i$) \cite{gutenkunst2009}, isolation with migration models \cite{hey2004}, and a series of method utilizing sequentially Markovian coalescent (PSMC, MSMC and SMC++) \cite{li2011, schiffels2014, terhorst2017}). 
Rather, we focus on methods that explicitly identify introgressed genomic segments based on patterns of nucleotide/haplotype diversity, level of differentiation and on phylogenetic relationships.
%We focus here on analytical prospects and limitations and introduce only a few representatives of each category.

@@

 \begin{equation}
 	RND_{min} = \frac{d_{min}}{d_{out}}
 \end{equation}
where $d_{min}$ is the minimum sequence distance between haplotypes in species X and Y; $d_{out}$ equals $(d_{XO} + d_{YO})/2$, the average sequence distance between each species and the outgroup ($O$).

These statistics have recently been further developed by adding differentiation between both non-admixed ($A$) and admixed populations ($B$) and a source population ($C$) \cite{racimo2016}. 
For example, the $U_{A,B,C(w,x,y)}$ statistic summarizes number of sites, where an allele at frequency $y$ in the source population ($C$) has a frequency higher than $x$ in the admixed population ($B$) and lower than $w$ in the non-admixed population ($C$).
A similar statistic, $Q95_{A,B,C(w,y)}$, sets a hard cutoff at the $95^{th}$ percentile of allele frequencies in the admixed population (B) \cite{racimo2016}.
\mbh{Did I get this right, Li?} \lwang{yes}
Further modifications have allowed specification of two source populations (see details in \cite{racimo2016}).

@@

Third, the ABBA-BABA statistic (also known as the D-statistic) and its derivatives are widely applied to introgression detection and make inferences based on genomic patterns of shared derived variants between populations or species in a phylogenetic context. \mbh{need sentence here explaining what D-statistic does} 
Elaborations of the D-statistic include $\hat{f_{d}}$ \cite{Martin2015} and the five-taxon D-statistic \cite{pease2015}. 
The former is quite similar to the D-statistic but uses allele frequencies from each population/species, and the latter detects introgression based on the localized phylogenetic pattern. \mbh{need to clarify here how this is different than $\hat{f_{d}}$}
%Their strength in identifying specific regions enables one to explore the relationships between introgression and recombination rate/gene density/distribution of deleterious alleles and to better understand the genomic regions penetrable to foreign gene flow.

Fourth, approaches that assign genome-wide genetic ancestry based on the popular software STRUCTURE \cite{pritchard2000} are another category to detect introgression. 
For example, fineSTRUCTURE \cite{Lawson2012} calculates the probability that genomic regions belong to genetic groups by \mbh{need details on method here}. 
GLOBETROTTER \cite{hellenthal2014} is a further development of fineSTRUCTURE that allows for identification of an unsampled source population and for dating of admixture events.

@@

Introgression appears to have played an important role in the domestication history of rice.
The center of Asian rice (\emph{Oryza sativa}) domestication is not known with complete confidence, but genetic and archaeobotanical evidence point toward both the Yangzee Basin in China and the Ganges plains in India. \mbh{there is a serious debate over whether there were one or two domestication centers of rice...I'd be sure to read up on the most current thoughts about this so that we don't upset potential reviewers}
Domestication occurred 8,200-13,500 BP from the wild species \emph{O. rufipogon} \cite{oka2012origin, fuller2010consilience, ricepedia, molina2011molecular}. \mbh{here limit to one or two citations}
Asian rice readily hybridizes with other domesticated subspecies and with wild relatives (of which there are about twenty \cite{ricepedia}). \mbh{probably best to find a primary literature citation for this (journal article rather than ricepedia).  What other domesticated subspecies are referred to here?  Also, when referring to Asian rice, is this both japonica and indica?  Finally, what are the twenty relatives?  My understanding is gene flow has mostly been with rufipogon and nivara}
The high genetic diversity within \emph{O. sativa} is likely due to introgression from wild relatives both within the domestication center(s) and in new environments where rice has dispersed following domestication \cite{second1982origin}.

Asian rice can be broadly clustered into two groups, japonica and indica \cite{khush2003classifying}.
%The vast majority of the rice varieties fell into the group indica (73.45\%) or japonica (23.02\%).
Japonica rice cultivars were likely domesticated first from wild rice populations in southern China, and indica cultivars later developed through hybridization of ancient japonica with new wild rice populations in south and southeast Asia \cite{Huang2012, londo2006phylogeography}. \mbh{Is this thought to be due to adaptation?  Did the wild rice populations confer local adaptation? If so, I'd state this and explain what type of adaptation this might have been if it's known}
%However, about 1.26\% belonged to four other groups (and 2.27\% were indeterminate).
%Some of the varieties from these other four groups most likely arose from hybrids of cultivated rice and \emph{O. rufipogon}.
Introgression of alleles between cultivated rice populations, as well as between cultivated and wild rice, is common \cite{oka2012origin, second1982origin, zhao2010genomic}\, and natural introgression from wild to domesticated rice is suspected (\cite{zhao2010genomic} calls for research into this possibility). \mbh{strange sentence here...says introgression between cultivated and wild rice is common but then says introgression from wild to cultivated is just suspected...need to clarify}
%These rices often have blight resistance gene, likely imparted from the wild relative parent.
Several resistance genes (grassy stunt virus, bacterial blight, brown planthopper, blast) are known to have been introgressed from wild relatives into \emph{O. sativa} by researchers \cite{brar1997alien, khush1974inheritance}. \mbh{don't think this is what we're after in these case studies.  We want to narrow our focus to natural introgression from wild to crop species during the expansion of the crop. Wild introgression during breeding is interesting, but not the focus of the paper}.
Some aus and rayada varieties retain characteristics of the putative wild parent. \mbh{more background on aus, rayada and aswina...what is there origin? how doe they relate to japonica and indica?}
Aus exhibits a sprawling growth pattern and easy-threshing grain.
Rayada is adapted to areas with longer periods of flooding, with heightened seed dormancy and photoperiod sensitivity (permitting harvest during times without standing water).
Aswina variety, adapted to growth in deepwater conditions with a short period of flooding, seems to be the result of hybridization between \emph{O. sativa} and deepwater-adapted \emph{O. rufipogon} populations. \mbh{need citations regarding these varieties}

In addition to investigative experiments, gene flow from wild relatives has been utilized to produce agronomic rice varieties.
Yatsen No. 1, for example, showed resistance to pests and diseases and adapted well to environmental conditions \cite{ting1933wild}.
Several lines were derived from Yatsen No. 1, and went on to be utilized extensively in parts of China. \mbh{again, don't think this is what we're after with these case studies}

>>>>

The process of domestication was once thought to be rapid and geographically constrained, with crops originating from a wild progenitor within one or more geographically defined centers followed by expansion to the modern-day extent of cultivation. \mbh{will need citation for this}
However, archaeological and genetic evidence are beginning to reveal that, in many cases, domestication has been temporally protracted and more diffuse \cite{brown2009complex, Meyer2016, wang2017, zhou2017}.
This new conception of domestication emphasizes the role of beneficial gene flow (\emph{i.e.}, adaptive introgression) from locally adapted wild relatives during crop expansion after initial domestication.

Adaptive introgression has three components: hybridization between two genomes, backcrossing to one of the parents, and selection on different recombinant genotypes with progressively diminished linkage drag \cite{barton2001role, Feuillet200824}.
In domesticated species, adaptive introgression would consist of crop/wild hybrids backcrossing to a crop, retention and increase in frequency of adaptive wild haplotypes in the crop, and removal of undesirable wild background.
To date, literature on crop-wild gene flow has focused on the risk of introgression of transgenes from domesticated crops into wild relatives (for a review, \cite{stewart2003transgene}) and on modern plant breeding efforts to introgress desired traits from wild relatives (for a review, \cite{zamir2001improving, tanksley1997seed, hajjar2007use}). \mbh{cite one paper here...the best, most recent and comprehensive review}
The history of natural introgression of wild alleles into domesticated crops over evolutionary timescales has received considerably less attention.
However, new tools and methods have recently been employed to detect genome-wide patterns of introgression, granting new insights into the prevalence of adaptive introgression in crop histories.
Emerging results suggest a need for reevaluation of the existing domestication paradigm.

@@

The decreasing cost of genome-wide resequencing and availability of reduced-representation genotyping (\emph{e.g.}, GBS and RAD-Seq), combined with new analytical methods, has facilitated comprehensive study of introgression across a number of species (\textbf{Table 1}).
High-density marker data can be used with haplotype-based and other methods to assign specific genomic regions to a taxon of origin and identify introgression across taxa \cite{Martin2015,Price2009,Lawson2012,pease2015,rosenzweig2016,geneva2015}.
The methods reviewed here do not include those marginally estimating introgression\slash migration rate as a component of demographic history (\emph{e.g.}, Approximate Bayesian Computation (ABC) \cite{beaumont2002}, diffusion approximations for demographic inference ($\delta a\delta i$) \cite{gutenkunst2009}, isolation with migration models \cite{hey2004}, and a series of methods utilizing the sequentially Markovian coalescent (PSMC, MSMC and SMC++) \cite{li2011, schiffels2014, terhorst2017}). 
Rather, we focus on methods that explicitly identify introgressed genomic segments based on patterns of nucleotide/haplotype diversity, the extent of differentiation, and phylogenetic relationships.
%We focus here on analytical prospects and limitations and introduce only a few representatives of each category.

@@

 \begin{equation}
 	RND_{min} = \frac{d_{min}}{d_{out}}
 \end{equation}
where $d_{min}$ is the minimum sequence distance between haplotypes in species X and Y and $d_{out}$ equals $(d_{XO} + d_{YO})/2$, the average sequence distance between each species and the outgroup ($O$).

These statistics have recently been further developed by adding differentiation between both non-admixed ($A$) and admixed populations ($B$) and a source population ($C$) \cite{racimo2016}. 
For example, the $U_{A,B,C(w,x,y)}$ statistic summarizes number of sites, where an allele at frequency $y$ in the source population ($C$) has a frequency higher than $x$ in the admixed population ($B$) and lower than $w$ in the non-admixed population ($C$).
A similar statistic, $Q95_{A,B,C(w,y)}$, sets a hard cutoff at the $95^{th}$ percentile of allele frequencies in the admixed population (B) \cite{racimo2016}.
Further modifications have allowed specification of two source populations (see details in \cite{racimo2016}).

@@

Third, the ABBA-BABA statistic (also known as the D-statistic) and its derivatives are widely applied to introgression detection.
These statistics make inferences regarding introgression based on genomic patterns of derived variants that are shared between populations or species.
Patterns of allele sharing are interpreted in a phylogenetic context and the method is best suited to detection of introgression at the genome level.
Elaborations of the D-statistic capable of localizing introgression to specific genomic regions include $\hat{f_{d}}$ \cite{Martin2015} and the five-taxon D-statistic \cite{pease2015}. 
The former is quite similar to the D-statistic but uses allele frequencies from each population/species, and the latter detects introgression based on the localized phylogenetic pattern.
%Their strength in identifying specific regions enables one to explore the relationships between introgression and recombination rate/gene density/distribution of deleterious alleles and to better understand the genomic regions penetrable to foreign gene flow.

The final category to detect introgression include approaches assigning genetic ancestry based on the popular software STRUCTURE \cite{pritchard2000}. 
For example, fineSTRUCTURE \cite{Lawson2012} calculates the probability that genomic regions belong to genetic groups by \mbh{need details on method here}. 
GLOBETROTTER \cite{hellenthal2014} is a further development of fineSTRUCTURE that allows for identification of an unsampled source population and for dating of admixture events.

@@

The story of the domestication of asian rice is still under debate.
A recent model utilizing molecular clocks points to single domestication of asian rice that occurred 8,200-13,500 BP in the Yangtze Basin in China from the wild species \emph{O. rufipogon}, with later divergence of the two prominent varietal groups indica and japonica lines \cite{molina2011molecular}.
Sequence data from indica, japonica, and wild O. rufipogon accessions points at japonica domestication in the Pearl Valley in China, and indica being formed from hybridization between japonica and rufipogon in Southern and South-eastern Asia \cite{Huang2012}.
On the other hand, genetic and archaeobotanical evidence points towards independent domestications of the indica and japonica subspecies in the Yangzee Basin and the Ganges plains in India \cite{fuller2010consilience}.

The authors of \cite{vaughan2008evolving} review the evidence for both hypotheses and support the single domestication event hypothesis, contributing the contrary evidence to the effects of a history of introgression, selection (primarily that of the non-shattering trait allele sh4), and diversification.
In their defense of the single-domestication hypothesis, the authors review the case that, during the spread of domesticated rice into new environments, gene flow from wild relatives would have imparted adaptive traits via introgression.
The wild relatives rufipogon and nivara are both phenotypically diverse and show evidence of local adaptation.
Rufipogon reproduction can be either primarily sexual or vegetative.
Whereas rufipogon is adapted to forested wetland environments, nivara is adapted to dryer conditions and has life cycle adaptations to survive grazing pressure.
Likewise, cultivated rice varieties display phenotypic differences important for fitness.
In particular, two of the cultivated rice deepwater varieties (rayada and ashwina) are said to be selected for the environment along the Ganges river.
Gene flow between these wild and cultivated rices is common, producing nuisance weedy hybrids in when the two are grown in sympatry.
However, during early domestication, introgression may have been even more prevalent because barriers to crop-to-wild introgression may have been less severe and because the inbreeding reproductive system of rice would not have been as firmly established.
Furthermore, the contemporary distribution of wild rice does not capture the range and diversity of wild rice during early domestication and range expansion of rice.
Introgression into indica is expected to be more likely than japonica, due to the higher degree of sympatry between indica and its wild relatives.
The high genetic diversity within domesticated rice is likely due to introgression from wild relatives both within the domestication center(s) and in new environments where rice has dispersed \cite{second1982origin}.
